---
title: "Modeling the presence/absence of species"
author: "Firza Riany"
date: "26/01/2021"
output: html_document
---

## To-do list

* Divide training and validation dataset  
* Train the model based on test and validation dataset  
* Create predicted Y values for training based on test and validation dataset  
* Evaluate model performance (RMSE between predicted vs observed for both test and validation dataset)  
* Check the assumption for logistic regression  
* In-depth EDA  

## Description

The project is aimed to understand the occurrences of tree species under
different temperature range. 

The information collected from this project can be useful to understand the
*distribution of tree species along elevational gradient* such as in mountain
forests. This is because elevational gradient is followed by temperature
gradient. 

Different temperatures along elevational gradient affect the distribution of
tree species because they require different temperature range to live optimally. 

## Disclaimer

This was a part of school project. Datasets were provided by my teacher from
AgroParisTech & Université de Lorraine, [Dr. Paulina Pinto](http://siafee.agroparistech.fr/pinto-paulina). I wrote the scripts under
her guidance. 

## Datasets

ecoflo dataset extracted from:  

* [Ecoplant](https://silvae.agroparistech.fr/home/?page_id=173), a database for soil and floristic data  
* Sophy  

Within the ecoflo dataset, there are:

* Floristic data that contains the presence/absence (1/0) of three tree species    
* Ecological data that contains:  
    + The elevation (alt)  
    + Optimal pH (VI pH)  
    + Soil condition (VI C/N)  
    + Average temperature in January (mt1), in July (mt7), and annual (mt13)  
    + Climatic water balance in July (bh7)  

```{r import file}
ecoflo = read.table("ecoflo.csv", header = TRUE, sep = ";")
```

Some remarks about the variables in the dataset:  

* Placette        : identifier for each tree observations 
* Periode         : if the observation is from old or new records    
* Massif          : the name of the mountains where the (tree) records were taken  
* xLamb and yLamb : geo-position of the plots where the records were taken   
* Annee           : the year the records were taken  
* Alt             : the latitude of the records  
* VIpH            : the optimal pH for the tree species  
* VICN            : the optimal soil condition for the tree species (CN means Carbon-Nitrogen ratio, it shows the amount of mineral contained in the soil. These are the organic materials that all tree species need to have)  
* Tree species    : there are three species in this dataset  
    + *Abies alba* (fir)  
    + *Larix decidua* (larch)  
    + *Quercus robur* (pedunculate oak)  
    + *Carpinus betulus* (Hornbeam)  
    + *Vaccinium myrtillus* (berry bush)  
* mt1             : average temperature in January  
* mt7             : average temperature in July  
* mt13            : annual average temperature  
* bh7             : climatic water balance in July  

## Exploratory Data Analysis (EDA)
```{r plot, fig.height=8, fig.width=10, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(hrbrthemes)
library(ggpubr)

# Abies presence
abies = ecoflo %>%
  filter(Abies_alba == 1)

abies_pres = ggplot(data = ecoflo, aes(x = xLamb2, y = yLamb2)) +
  geom_point(shape = 1, alpha = 0.4, size = 1.5, color = "gray70") +
  geom_point(data = abies, aes(x = xLamb2, y = yLamb2),
             color = "indianred3", size = 1.5, alpha = 0.4,
             shape = 1) +
  labs(subtitle = "Abies alba presence") +
  xlab("Coordinate X") + ylab("Coordinate Y") +
  theme_ipsum(base_size = 7)

# Carpinus presence
carpinus = ecoflo %>%
  filter(Carpinus_betulus == 1)

carp_pres = ggplot(data = ecoflo, aes(x = xLamb2, y = yLamb2)) +
  geom_point(shape = 1, alpha = 0.4, size = 1.5, color = "gray70") +
  geom_point(data = carpinus, aes(x = xLamb2, y = yLamb2),
             color = "cyan4", size = 1.5, alpha = 0.4,
             shape = 1) +
  labs(subtitle = "Carpinus betulus presence") +
  xlab("Coordinate X") + ylab("Coordinate Y") +
  theme_ipsum(base_size = 7)

# Larix presence
larix = ecoflo %>% 
  filter(Larix_decidua == 1)

larix_pres = ggplot(data = ecoflo, aes(x = xLamb2, y = yLamb2)) +
  geom_point(shape = 1, alpha = 0.4, size = 1.5, color = "gray70") +
  geom_point(data = larix, aes(x = xLamb2, y = yLamb2),
             color = "dark orange2", size = 1.5, alpha = 0.4,
             shape = 1) +
  labs(subtitle = "Larix decidua presence") +
  xlab("Coordinate X") + ylab("Coordinate Y") +
  theme_ipsum(base_size = 7)

# Vacciniul presence
vacc = ecoflo %>%
  filter(Vaccinium_myrtillus == 1)

vacc_pres = ggplot(data = ecoflo, aes(x = xLamb2, y = yLamb2)) +
  geom_point(shape = 1, alpha = 0.4, size = 1.5, color = "gray70") +
  geom_point(data = vacc, aes(x = xLamb2, y = yLamb2),
             color = "darkviolet", size = 1.5, alpha = 0.4,
             shape = 1) +
  labs(subtitle = "Vaccinium myrtillus presence") +
  xlab("Coordinate X") + ylab("Coordinate Y") +
  theme_ipsum(base_size = 7)

presence = ggarrange(abies_pres, carp_pres, larix_pres, vacc_pres)

presence
```

```{r summary}
ecoflo$mt13[is.na(ecoflo$mt13)] = 9.5
summary(ecoflo)
ecoflo$massif = as.factor(ecoflo$massif)
ecoflo$periode = as.factor(ecoflo$periode)
ecoflo$annee = as.factor(ecoflo$annee)
```

So, my research question is: "How is the presence of tree species affected by
temperature gradient?"

There is a theory that temperature is changing alongside altitudinal gradient. So,
if we know the altitude of a tree species, we can use the altitude as the position of
the tree and relate it with temperature. We can see how temperature can affect the
presence of trees along altitudinal gradient.

```{r altitude hist, message=FALSE, warning=FALSE}
# Temperature distribution
hist_alt = ggplot(data = ecoflo, aes(x = mt13)) +
  geom_histogram(bins = 20, fill = "cyan4", color = "gray92", alpha = 0.8) +
  geom_vline(aes(xintercept = mean(mt13), color = "Mean"), size = 1) +
  geom_vline(aes(xintercept = median(mt13), color = "Median"), size = 1) +
  scale_color_manual(name = "Statistics", values = c("Mean" = "red", 
                                                     "Median" = "orange")) +
  labs(title = "Temperature distribution") +
  xlab("Temperature (°Celcius)") +
  ylab("Counts") +
  theme_ipsum(plot_title_size = 10)

hist_alt
```

```{r species presence per temperature, message=FALSE, warning=FALSE}
# Temperature distribution for the presence of Abies alba
abies_temp = ggplot(data = abies, aes(x = mt13)) +
  geom_histogram(bins = 20, fill = "indianred3", color = "gray92", alpha = 0.8) +
  geom_vline(aes(xintercept = mean(mt13), color = "Mean"), size = 1) +
  geom_vline(aes(xintercept = median(mt13), color = "Median"), size = 1) +
  scale_color_manual(name = "Statistics", values = c("Mean" = "red", 
                                                     "Median" = "orange")) +
  labs(title = "Temperature distribution for Abies alba") +
  xlab("Temperature (°Celcius)") +
  ylab("Counts") +
  xlim(0, 20) +
  ylim(0, 1000) +
  theme_ipsum(plot_title_size = 10, plot_title_margin = 4) +
  theme(legend.position = "none")

# Temperature distribution for the presence of Carpinus betulus
carp_temp = ggplot(data = carpinus, aes(x = mt13)) +
  geom_histogram(bins = 20, fill = "cyan4", color = "gray92", alpha = 0.8) +
  geom_vline(aes(xintercept = mean(mt13), color = "Mean"), size = 1) +
  geom_vline(aes(xintercept = median(mt13), color = "Median"), size = 1) +
  scale_color_manual(name = "Statistics", values = c("Mean" = "red", 
                                                     "Median" = "orange")) +
  labs(title = "Carpinus betulus") +
  xlab("Temperature (°Celcius)") +
  ylab("Counts") +
  xlim(0, 20) +
  ylim(0, 1000) +
  theme_ipsum(plot_title_size = 10, plot_title_margin = 4)

# Temperature distribution for the presence of Larix decidua
larix_temp = ggplot(data = larix, aes(x = mt13)) +
  geom_histogram(bins = 20, fill = "dark orange2", color = "gray92", alpha = 0.8) +
  geom_vline(aes(xintercept = mean(mt13), color = "Mean"), size = 1) +
  geom_vline(aes(xintercept = median(mt13), color = "Median"), size = 1) +
  scale_color_manual(name = "Statistics", values = c("Mean" = "red", 
                                                     "Median" = "orange")) +
  labs(title = "Larix decidua") +
  xlab("Temperature (°Celcius)") +
  ylab("Counts") +
  xlim(0, 20) +
  ylim(0, 1000) +
  theme_ipsum(plot_title_size = 10, plot_title_margin = 4) +
  theme(legend.position = "none")

# Temperature distribution for the presence of Vaccinium myrtillus
vacc_temp = ggplot(data = vacc, aes(x = mt13)) +
  geom_histogram(bins = 20, fill = "darkviolet", color = "gray92", alpha = 0.8) +
  geom_vline(aes(xintercept = mean(mt13), color = "Mean"), size = 1) +
  geom_vline(aes(xintercept = median(mt13), color = "Median"), size = 1) +
  scale_color_manual(name = "Statistics", values = c("Mean" = "red", 
                                                     "Median" = "orange")) +
  labs(title = "Vaccinium myrtillus") +
  xlab("Temperature (°Celcius)") +
  ylab("Counts") +
  xlim(0, 20) +
  ylim(0, 1000) +
  theme_ipsum(plot_title_size = 10, plot_title_margin = 4)

temp_dist = ggarrange(abies_temp, carp_temp, larix_temp, vacc_temp)

temp_dist
```

But, how is the relationship between altitude and temperature? Are they really
negatively or positively related? To know this, we should run a simple linear
model.

```{r linear alt x temp, message=FALSE, warning=FALSE}
# Relationship between altitude and mean annual temperature
lm_temp = lm(ecoflo$mt13 ~ ecoflo$alt)

summary(lm_temp) # an increase in alt leads to a decrease in temperature

temp_plot = ggplot(data = ecoflo, aes(x = alt, y = mt13)) +
  geom_point(shape = 1, size = 1.5, color = "violetred4", alpha = 0.4) +
  geom_smooth(method = "lm", na.rm = TRUE, color = "dodgerblue3", alpha = 0.5,
              lwd = 1) +
  labs(title = "Relationship between temperature and altitude") +
  xlab("Altitude (meter above sea level)") +
  ylab("Mean annual temperature (°Celcius)") +
  theme_ipsum(plot_title_size = 12, base_size = 10, plot_title_margin = 4)
```

Relationship between altitude and temperature is negative relationship. It means 
an increase in altitude will cause a decrease in temperature. It also means that
both variables are related, so we can either use temperature or altitude to predict
the presence of a tree species. 

In this project I will see the ecological profile of Abies alba using temperature. 

```{r logit, message=FALSE, warning=FALSE}
# Presence of Abies alba as a function of temperature
abies_glm = glm(ecoflo$Abies_alba ~ ecoflo$mt13, family = binomial)

summary(abies_glm)
# 9565.3 - 8708.3, deviation = 857 --> the variance explained by temperature

# The significance of the model with temperature
anova(abies_glm, test="Chisq") # temperature can explain the presence of abies alba

# Predicted values = abies_temp$fitted.values

# Storing the fitted values from abies_temp in a dataset
predicted = data.frame(fitted(abies_glm))

names(predicted)[1] = "Abies_alba"

# Presence of Carpinus betulus as a function of temperature
carp_glm = glm(ecoflo$Carpinus_betulus ~ ecoflo$mt13, family = binomial)

summary(carp_glm)
# 4461.0 - 4457.9 

anova(carp_glm, test="Chisq") # temperature is not significant

# Presence of Larix decidua
larix_glm = glm(ecoflo$Larix_decidua ~ ecoflo$mt13, family = binomial)

summary(larix_glm)
# 3565.8 - 2557.8, deviance = 1008 --> variance explained by temperature

anova(larix_glm, test="Chisq") # temperature has a significant effect on the presence

predicted$Larix_decidua = fitted(larix_glm)

# Presence of Vaccinium
vacc_glm = glm(ecoflo$Vaccinium_myrtillus ~ ecoflo$mt13, family = binomial)

summary(vacc_glm)
# 7988.7  - 6773.3, deviance = 1200 --> variance explained by temperature

anova(vacc_glm, test="Chisq") # temperature has a significant effect

predicted$Vaccinium_myrtillus = fitted(vacc_glm)

library(tidyr)

predicted_long = gather(data = predicted, key = species, value = glm_predict,
                        Abies_alba:Vaccinium_myrtillus, factor_key = TRUE)

# temperature repeated 3 times, 1 - 7695
predicted_long$temp = rep(ecoflo$mt13, times = 3)

summary(predicted_long$species)

fit_2 = ggplot(data = predicted_long, 
               aes(x = temp, y = glm_predict, color = species)) +
  geom_point(aes(color = species),
             alpha = 0.7, size = 1.5) +
  labs(title = "Probability of tree presence across temperature gradients") +
  xlab("Temperature (°Celcius)") +
  ylab("Predicted probabilities") +
  scale_color_manual(name = "Species",
                     values = c("Abies_alba" = "indianred3",
                                "Larix_decidua" = "dark orange2",
                                "Vaccinium_myrtillus" = "darkviolet")) +
  theme_ipsum(plot_title_size = 12, plot_title_margin = 4)
```